Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.qmd
Update paper

The scope of a kinetic model is defined by a stoichiometric matrix, $S$, in
which rows represent metabolites, columns represent reactions, and matrix
elements $s_{ij}$ represent the stoichiometric coefficient of metabolite $i$ in
reaction $j$. The change in metabolite concentrations is

@@

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in our case study below, capturing this correlation structure is difficult
outside a Bayesian context.

@@

After specifying the qualitative aspects of the kinetic model, we chose
parameter values to use as ground truth by Monte Carlo sampling using a previous
model of the methionine cycle as a starting point (see @saa_general_2015 for
this model).

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
data datasets, obtaining XXX post-warmup samples after XXX of total computation
time. For the Laplace approximation comparison, we used Maud’s Laplace mode.
Full details as well as instructions for reproducing our analysis can be found
at XXX.

@@

Running standard diagnostic checks indicated that the samples we generated
were from the target posterior distribution. The improved $\hat{R}$ statistic
[@vehtariRankNormalizationFoldingLocalization2021] for every sampled variable
was within 1% of 1, indicating appropriate mixing within and between Markov
chains. Additionally, the number of effective samples was high, indicating
that we generated enough posterior samples to support inferences about the
bulks of the distributions of the sampled parameters. Furthermore, we observed
no post-warmup divergent transitions, indicating that the sampler was able to
transform the log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such as the
one shown in the bottom left of figure 3; in this case the marginal distribution
of the two parameters is roughly banana shaped. More commonly, however, two-
dimensional pair plots were insufficient to reveal the underlying correlation
structure, as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations involve more
than two parameters.

@@

Our results here also provide circumstantial evidence that the parameter
correlations in Bayesian kinetic model posteriors tend to be non-linear, as
a posterior with only linear correlations would likely be more germane to
Laplace approximation.A conclusion that we drew from this analysis was that
the results of fitting our model cannot be summarised simply, for example by
fitting a multivariate normal distribution to the posterior draws. We therefore
recommend that Maud users store the full set of MCMC draws rather than using
such an approximation. This does not preclude the possibility that there is
an alternative, more compact, way to summarise the results of Bayesian kinetic
model inference; we leave research into this topic to future work.

@@

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, Macos
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud- metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

@@

We chose to implement a custom input format despite the existence of standard
formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features of
SBML and PEtab that make them structurally unsuitable in this context. Our
requirements for an input format included that it be mathematics-free, so that
all mathematical details are encapsulated in source code, and that it has a
detailed, verifiable structure. These requirements made toml more attractive
than SBML: toml is easier for humans to read and edit and can straightforwardly
be validated using tools like pydantic. Further, an SBML representation of our
desired input would not contain differential equations.  It would therefore not
be interoperable with most SBML-targeting software, which typically assumes that
differential equations are available and does not know about Maud’s structure.

@@

Except for metabolites’ standard condition Gibbs energy changes of formation,
Maud uses independent normal prior distributions for parameters that can in
principle be both negative and positive. For parameters that are constrained to
be positive, Maud uses independent log-normal distributions. Formation energy
parameters have a multivariate normal prior distribution. Location, scale and
covariance parameters for all these prior distributions can be chosen freely by
the user.

@@

All measurements are specific to an experimental condition - that is, a case
where the true state of the network, including knockouts, boundary conditions
and state variables as well as kinetic and thermodynamic parameters, can safely
be assumed to be the same. Maud’s statistical model allows for arbitrarily many
experimental conditions, and for any measurable quantity to be measured any
number of times in any condition.

@@

Flux measurements, representing the results of quantitative fluxomics analyses,
are modelled using a likelihood function from a standard linear regression
model, as shown in equation Equation 8 (REFERENCE). Flux measurements can be
obtained from the results of isotope labelling experiments using metabolic
flux analysis, for example as described in (Young 2014). When entering flux
measurements, it is important only to specify measurements for a network’s free
fluxes, as the values of some steady state fluxes in a metabolic network are
constrained by others, with the result that dependent fluxes cannot typically be
measured separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

@@

Finally, the reader may wonder why we chose a linear regression model for
reaction flux measurements even though this creates the potential for erroneous
double counting and requires non-trivial upstream modelling, as intracellular
fluxes typically cannot be measured directly, but must be inferred from isotope
labelling experiments using metabolic flux analysis (see REFERENCE). Ideally
Maud’s measurement model for fluxes would extend from fluxes to the results of
potential labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued, again
for the sake of simplicity.

@@

Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
to implement a command line interface. The command line interface loads
input files as Python dictionaries, which are parsed using the Python library
`toml` [@pearsonTomlPythonLibrary2020] and then validated and converted into
structured `MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022].
Maud’s statistical model is implemented in the probabilistic programming
language Stan [@carpenterStanProbabilisticProgramming2017] and accessed using
the interface cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior
sampling, Maud uses the `MaudInput` to create a Stan input json file and
obtain configuration information for cmdstanpy, which it uses to (if necessary)
generate a model executable file and then trigger posterior sampling using
adaptive Hamiltonian Monte Carlo. When sampling is complete, Maud converts to
the output into the standard format InferenceData using the Python library arviz
[@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
information for debugging.

@@

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
accessed via their interface from Stan. The hybrid method follows that propsed
by @margossianComputingSteadyStates2018, and involves numerically evolving
the ODE system for a short period of time, then using the difference between
the evolved and starting concentrations as the target for a numerical algebra
solver.

>>>>

The scope of a kinetic model is defined by a stoichiometric matrix, $S$, in
which rows represent metabolites, columns represent reactions, and matrix
elements $s_{ij}$ represent the stoichiometric coefficient of metabolite $i$ in
reaction $j$. The change in metabolite concentrations is:

@@

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in our case study below, capturing this correlation structure is difficult
outside a Bayesian context.

@@

After specifying the qualitative aspects of the kinetic model, we selected
parameter values to use as ground truth by Monte Carlo sampling using a previous
model of the methionine cycle as a starting point (see @saa_general_2015 for
this model).

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
-data datasets, obtaining XXX post-warmup samples after XXX of total computation
time. For the Laplace approximation comparison, we used Maud’s Laplace mode.
Full details as well as instructions for reproducing our analysis can be found
at XXX.

@@

Running standard diagnostic checks indicated that the samples we generated
were from the target posterior distribution. The improved $\hat{R}$ statistic
[@vehtariRankNormalizationFoldingLocalization2021] for every sampled variable
was within 1% of 1, indicating appropriate mixing within and between Markov
chains. Additionally, the number of effective samples was high, indicating
that we generated enough posterior samples to support inferences about the
bulks of the distributions of the sampled parameters. Furthermore, we observed
no post warm-up divergent transitions, indicating that the sampler was able to
transform the log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such
as the one shown in the bottom left of figure 3; in this case the marginal
distribution of the two parameters is roughly banana shaped. More commonly,
however, two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure 3. This does
not mean that the parameters were uncorrelated, but rather that the correlations
involve more than two parameters.

@@

Our results here also provide circumstantial evidence that the parameter
correlations in Bayesian kinetic model posteriors tend to be non-linear, as
a posterior with only linear correlations would likely be more germane to
Laplace approximation. A conclusion that we drew from this analysis was that
the results of fitting our model cannot be summarised simply, for example by
fitting a multivariate normal distribution to the posterior draws. We therefore
recommend that Maud users store the full set of MCMC draws rather than using
such an approximation. This does not preclude the possibility that there is
an alternative, more compact, way to summarise the results of Bayesian kinetic
model inference; we leave research into this topic to future work.

@@

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud- metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

@@

We chose to implement a custom input format despite the existence of standard
formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features of
SBML and PEtab that make them structurally unsuitable in this context. Our
requirements for an input format included that it be mathematics-free, so that
all mathematical details are encapsulated in source code, and that it has a
detailed, verifiable structure. These requirements made toml more attractive
than SBML: toml is easier for humans to read and edit and can straightforwardly
be validated using tools like Pydantic. Further, an SBML representation of our
desired input would not contain differential equations.  It would therefore not
be interoperable with most SBML targeting software, which typically assumes that
differential equations are available and does not know about Maud’s structure.

@@

Except for metabolites’ standard condition Gibbs energy changes of formation,
Maud uses independent normal prior distributions for parameters that can in
principle be both negative and positive. For parameters that are constrained to
be positive, Maud uses independent log-normal distributions. Formation energy
parameters have a multivariate normal prior distribution. Location, scale and
covariance parameters for all these prior distributions can be selected freely
by the user.

@@

All measurements are specific to an experimental condition; that is, a case
where the true state of the network, including knockouts, boundary conditions
and state variables as well as kinetic and thermodynamic parameters, can safely
be assumed to be the same. Maud’s statistical model allows for arbitrarily many
experimental conditions, and for any measurable quantity to be measured any
number of times in any condition.

@@

Flux measurements, representing the results of quantitative fluxomics analyses,
are modelled using a likelihood function from a standard linear regression
model, as shown in equation 8 (REFERENCE). Flux measurements can be obtained
from the results of isotope labelling experiments using metabolic flux analysis,
for example as described in (Young 2014). When entering flux measurements,
it is important only to specify measurements for a network’s free fluxes, as
the values of some steady state fluxes in a metabolic network are constrained
by others, with the result that dependent fluxes cannot typically be measured
separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

@@

Finally, the reader may wonder why Maud uses a linear regression model for
reaction flux measurements even though this creates the potential for erroneous
double counting and requires non-trivial upstream modelling, as intracellular
fluxes typically cannot be measured directly, but must be inferred from isotope
labelling experiments using metabolic flux analysis (see REFERENCE). Ideally
Maud’s measurement model for fluxes would extend from fluxes to the results of
potential labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued, again
for the sake of simplicity.

@@

Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
to implement a command line interface. The command line interface loads input
files as Python dictionaries, which are parsed using the Python library `toml`
[@pearsonTomlPythonLibrary2020] and then validated and converted into structured
`MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022]. Maud’s
statistical model is implemented in the probabilistic programming language Stan
[@carpenterStanProbabilisticProgramming2017] and accessed using the interface
cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior sampling, Maud
uses the `MaudInput` to create a Stan input json file and obtain configuration
information for cmdstanpy, which it uses to (if necessary) generate a
model executable file and then trigger posterior sampling using adaptive
Hamiltonian Monte Carlo. When sampling is complete, Maud converts to the
output into the standard format `InferenceData` using the Python library arviz
[@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
information for debugging.

@@

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
accessed via their interface from Stan. The hybrid method follows that proposed
by @margossianComputingSteadyStates2018, and involves numerically evolving
the ODE system for a short period of time, then using the difference between
the evolved and starting concentrations as the target for a numerical algebra
solver.
